To be FAIR: Theory Specification Needs an Update

### 2025-03-06 manuscript.Rmd

The so-called "replication crisis" has prompted extensive scientific reforms [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
Concern that the abundance of non-replicable findings 
was caused by undisclosed flexibility in analyses led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
These various practices ensure transparent and repeated testing of hypotheses by commiting to an analysis plan in advance.
However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].
Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
Psychological theories rarely provide hypotheses that are corroborated by evidence.
Furthermore, theories are often so ambiguous that they can accommodate mutually inconsistent findings,
as the theory's central claims evade falsification.
A good example of this is found in “self-determination theory” (SDT), which emphasizes the role of intrinsic and extrinsic motivation in human behavior.
Initially, intrinsic motivation was understood as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures [@deciEffectsExternallyMediated1971].
Over time, however, SDT expanded its definition to include motivations driven by the fulfillment of basic psychological needs—autonomy, competence, and relatedness—all still categorized as "intrinsic" [@ryanSelfdeterminationTheoryFacilitation2000].
The difference between these definitions becomes clear when considering the act of changing an infant's dirty diaper.
Under the original definition, one might hypothesize that caregivers are not intrinsically motivated to change diapers, as this is hardly a joyous experience.
Under the expanded definition, one might hypothesize the opposite, as the act may fulfill the need for relatedness.

>>>>

The so-called "replication crisis" has prompted extensive scientific reforms [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
Concern that the abundance of non-replicable findings 
was caused by undisclosed flexibility in analyses led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
These various practices ensure transparent and repeated testing of hypotheses by commiting to an analysis plan in advance.
However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].
Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:

Psychological theories rarely provide hypotheses that are corroborated by evidence.
Furthermore, theories are often so ambiguous that they can accommodate mutually inconsistent findings,
as the theory's central claims evade falsification.
A good example of this is found in “self-determination theory” (SDT), which emphasizes the role of intrinsic and extrinsic motivation in human behavior.
Initially, intrinsic motivation was understood as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures [@deciEffectsExternallyMediated1971].
Over time, however, SDT expanded its definition to include motivations driven by the fulfillment of basic psychological needs—autonomy, competence, and relatedness—all still categorized as "intrinsic" [@ryanSelfdeterminationTheoryFacilitation2000].
The difference between these definitions becomes clear when considering the act of changing an infant's dirty diaper.
Under the original definition, one might hypothesize that caregivers are not intrinsically motivated to change diapers, as this is hardly a joyous experience.
Under the expanded definition, one might hypothesize the opposite, as the act may fulfill the need for relatedness.
